### 2020-04-14 manuscript/manuscript.rmd
renamed code blocks, moved some text around

# Rooting the animal tree of life - more is less for Ctenophora-sister?

@@

## Abstract (ideally less than 200 words)
(Current version)
Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships in the animal tree of life. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implications for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We next quantify the support of phylogenetic signal and conduct sensitive analyses in several representative data matrices. We found that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions. We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic. Thus, our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

(Casey's edit)
Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implciations for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from a large number of previous phylogenomic analyses of protein coding genes. These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used. As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes. Most other conditions supported Ctenophora-sister. It has been argued that the results using site heterogeneous models with reduced outgroup sampling (which favor Porifera-Sister) should be favored because they best address different frequencies of amino acids found across sites and in more distantly related outgroups. We perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We performed additional analyses using new site heterogeneous models in iqtree, and find that, unlike the phylobayes analyes, these support Ctenophora-sister rather than Porifera-sister. This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used.
We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models). We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic. Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

## Main

@@

```{r overview data matrices}
tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")

@@

```{r taxon_rectangles, fig.height = 18, fig.width = 20 }

@@

```{r gene_composition,, fig.height = 6, fig.width = 10}
matrix_summary = merge(matrix_summary,n_ribo,by="matrix")

```{r matrix_overlap, fig.height = 12, fig.width = 20}

@@

```{r support_published_analyses}
analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
x1 = factor(analyses_published$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
analyses_published %>%
	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
   xlab("Model complexity") + ylab("Clade")
```
**Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Supplementary Table 2).

@@

```{r support_new_analyses}
analyses_new=analyses_new %>%
mutate(model=recode(model,
                    `GTR20`="GTR",
                    `CAT+F81`="Poisson+CAT"))

x1 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
analyses_new %>%
	ggplot( aes(x=x1, y=clade, col=result, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
       xlab("Model complexity") + ylab("Clade")
```
**Fig. 4.**  A total of `r nrow(analyses_new)` new phylgoenomic analyses were conducted in this study (Supplementary Table 3).

@@

**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
```{r modelfinder}
analyses_new %>%
	filter(modelfinder==TRUE) %>%
	select( matrix, clade, result, model_summary ) %>%
	kable()

@@

```{r phylogenetic_signal, fig.width=6, fig.height=12}
# fig5_bars <- au_tests %>%
au_tests %>%
  unite(Matrix, c(manuscript, matrix, model)) %>%
  group_by(Matrix, result) %>%
  count() %>%
  group_by(Matrix) %>%
  mutate(percent = prop.table(n)) %>%

@@

## Future direction
Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised. Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution. First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available). Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated]. Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem. More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work. Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian]. We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.

@@

# Supplemental Information

@@

## Supplementary Tables
```{r phylogenetic signal distribution}
tableXXdistribution = read_tsv("../trees_new/AU_test/summary.tsv")
	kable(tableXXdistribution, caption= "Supplementary Table 4. Distribution of phylogenetic signal for different models and outgroup choice")

@@

>>>>

# Rooting the animal tree of life

@@

## Abstract

Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implciations for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from a large number of previous phylogenomic analyses of protein coding genes. These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used. As other authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes. Most other conditions supported Ctenophora-sister. It has been argued that the results using site heterogeneous models with reduced outgroup sampling (which favor Porifera-Sister) should be favored because they best address different frequencies of amino acids found across sites and in more distantly related outgroups. We perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We performed additional analyses using new site heterogeneous models in iqtree, and find that, unlike the phylobayes analyes, these support Ctenophora-sister rather than Porifera-sister. This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used. We also quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models). We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic. Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

## Introduction

@@

@@

```{r figure_taxon_rectangles, fig.height = 18, fig.width = 20 }

@@

@@

```{r figure_support_analyses_published}
analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
x1 = factor(analyses_published$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
analyses_published %>%
	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
   xlab("Model complexity") + ylab("Clade")
```
**Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table XX).

@@

```{r figure_support_analyses_new}
analyses_new=analyses_new %>%
mutate(model=recode(model,
                    `GTR20`="GTR",
                    `CAT+F81`="Poisson+CAT"))

x1 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
analyses_new %>%
	ggplot( aes(x=x1, y=clade, col=result, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
       xlab("Model complexity") + ylab("Clade")
```
**Fig. 4.**  A total of `r nrow(analyses_new)` new phylgoenomic analyses were conducted in this study (Supplementary Table XX).

@@

@@

```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
# fig5_bars <- au_tests %>%
au_tests %>%
  unite(Matrix, c(manuscript, matrix, model)) %>%
  group_by(Matrix, result) %>%
  count() %>%
  group_by(Matrix) %>%
  mutate(percent = prop.table(n)) %>%

@@

Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised. Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution. First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available). Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated]. Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem. More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work. Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian]. We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.

@@

# Supplementary Information

@@

## Extended Tables

```{r table_study_summary}
tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")

@@

```{r table_phylogenetic_signal}
tableXXdistribution = read_tsv("../trees_new/AU_test/summary.tsv")
	kable(tableXXdistribution, caption= "Supplementary Table 4. Distribution of phylogenetic signal for different models and outgroup choice")

**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
```{r table_modelfinder}
analyses_new %>%
	filter(modelfinder==TRUE) %>%
	select( matrix, clade, result, model_summary ) %>%
	kable()

## Extended Figures

```{r figure_gene_composition,, fig.height = 6, fig.width = 10}
matrix_summary = merge(matrix_summary,n_ribo,by="matrix")

```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}

## Supplementary Information
